Piecewise Monotone Quintic Spline Interpolants

### 2020-06-12 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/old/pmqsi.tex
Working through a new version of MQSI. Introduced some kind of Heisenbug, still investigating.

%**************************************************************************
%* SpringSim 2020 Author Kit
%*
%* Word Processing System: TeXnicCenter and MiKTeX
%*
%**************************************************************************

\documentclass{scspaperproc}

\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{mathptmx}

%
%****************************************************************************
% AUTHOR: You may want to use some of these packages. (Optional)
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsbsy}
\usepackage{amsthm}
%****************************************************************************

%
%****************************************************************************
% AUTHOR: If you do not wish to use hyperlinks, then just comment
% out the hyperref usepackage commands below.

%% This version of the command is used if you use pdflatex. In this case you
%% cannot use ps or eps files for graphics, but pdf, jpeg, png etc are fine.

\usepackage[pdftex,colorlinks=true,urlcolor=blue,citecolor=black,anchorcolor=black,linkcolor=black,bookmarks=false]{hyperref}

%% The next versions of the hyperref command are used if you adopt the
%% outdated latex-dvips-ps2pdf route in generating your pdf file. In
%% this case you can use ps or eps files for graphics, but not pdf, jpeg, png etc.
%% However, the final pdf file should embed all fonts required which means that you have to use file
%% formats which can embed fonts. Please note that the final PDF file will not be generated on your computer!
%% If you are using WinEdt or PCTeX, then use the following. If you are using
%% Y&Y TeX then replace "dvips" with "dvipsone"

%% \usepackage[dvips,colorlinks=true,urlcolor=blue,citecolor=black,%
%% anchorcolor=black,linkcolor=black]{hyperref}

%% The use of the long citation format (e.g. "Brown and Edwards (1993)" rather than "[5]") and at the same
%% time using the hyperref package can lead to hard to trace bugs in case the citation is broken accross the
%% line (usually this will mark the entire paragraph as a hyperlink (clickable) which is easily noticeable and fixed
%% if using colorlinks, but not if the color is black -- as it is now). Worse yet, if a citation spans page boundary,
%% LaTeX compilation can fail, with an obscure error message. Since this depends a lot on the flow of the text
%% and wording, these bugs come and go and can be extremely hard for a beginner to trace. The error
%% message can look like this:
%%
%%    ! pdfTeX error (ext4): \pdfendlink ended up in different nesting level than \pdfstartlink.
%%    \AtBegShi@Output ...ipout \box \AtBeginShipoutBox 
%%    \fi \fi 
%%    l.174 
%%    ! ==> Fatal error occurred, no output PDF file produced!
%%
%% and can be universally fixed by putting an \mbox{} around the citation in question (in this case, at line 174)
%% and maybe adapting the wording a little bit to improve the paragraph typesetting, which is perhaps not
%% immediately obvious.
%****************************************************************************

%****************************************************************************
%*
%* AUTHOR: YOUR CALL!  Document-specific macros can come here.
%*
%****************************************************************************

%% Use a package to allow for multi-row cells in a table.
\usepackage{multirow}

%% Use a custom `algorithm` command for this paper.
\input{lux-algorithm}

% add custom hyphenation rules here
\usepackage{hyphenat}
\hyphenation{op-tical net-works semi-conduc-tor}

% If you use theorems
\newtheoremstyle{scsthe}% hnamei
{8pt}% hSpace abovei
{8pt}% hSpace belowi
{\it}% hBody fonti
{}% hIndent amounti1
{\bf}% hTheorem head fontbf
{.}% hPunctuation after theorem headi
{.5em}% hSpace after theorem headi2
{}% hTheorem head spec (can be left empty, meaning `normal')i

\theoremstyle{scsthe}
\newtheorem{theorem}{Theorem}
\renewcommand{\thetheorem}{\arabic{theorem}}
\newtheorem{corollary}[theorem]{Corollary}
\renewcommand{\thecorollary}{\arabic{corollary}}
\newtheorem{definition}{Definition}
\renewcommand{\thedefinition}{\arabic{definition}}

% avoid overrunning the right margin; you are welcome to remove this, provided that you take care not to overrun the right margin anywhere in your paper
\sloppy

%#########################################################
%*
%*  The Document.
%*
\begin{document}

%***************************************************************************
% AUTHOR: Thomas C.H. Lux et al.
% FORMAT AUTHORS NAMES Like: Author1, Author2 and Author3 (last names)
%
%		You need to change the author listing below!
%               Please list ALL authors using last name only, separate by a comma except
%               for the last author, separate with "and"

\SCSpagesetup{Lux, Watson, Chang, Xu, Wang, and Hong}

% AUTHOR: Uncomment ONE of these correct conference names.
\def\SCSconferenceacro{SpringSim'20}
%\def\SCSconferenceacro{SummerSim}
%\def\SCSconferenceacro{AutumnSim}
%\def\SCSconferenceacro{PowerPlantSim}

% AUTHOR: Set the correct year of the conference.
\def\SCSpublicationyear{2020}

% AUTHOR: Set the correct month and dates; the dates are separated by a single minus sign
% with no spaces and no leading zeros, the month is a full name (e.g. April) with the first letter
% capitalized. For example, "April 8-13".
\def\SCSconferencedates{May 19-May 21}

% AUTHOR: Set the correct venue in the form "City, State, Country", for example "Los Angeles, CA, USA".
\def\SCSconferencevenue{Fairfax, VA, USA}

% AUTHOR: Enter the title, all letters in upper case
\title{An Algorithm for Constructing Monotone \\ Quintic Interpolating Splines}

% AUTHOR: Enter the authors of the article, see end of the example document for further examples
\author{
Thomas C. H. Lux\\
Layne T. Watson\\
Tyler H. Chang\\ [12pt]
Departments of Computer Science, Mathematics,\\
and Aerospace and Ocean Engineering\\
Virginia Polytechnic Institute \& State University\\
2000 Torgersen Hall, Blacksburg, VA, USA\\
\{tchlux,ltw,thchang\}@vt.edu\\
\and
Li Xu\\%
Yueyao Wang\\%
Yili Hong\\ [12pt]
Department of Statistics\\
Virginia Polytechnic Institute \& State University\\
213 Hutcheson Hall, Blacksburg, VA, USA\\
\{lix1992,yueyao94,yilihong\}@vt.edu\\
}

\maketitle

%% Examples and notes.
%% 
%% \begin{table}[htb]
%%   \centering
%%   \caption{Caption above the table!}\label{tab:example}
%%   \begin{tabular}{r|l}
%%     First & row \\ \hline
%%     second & row \\
%%     third & row \\
%%   \end{tabular}
%% \end{table}
%% 
%% \begin{figure}[htb]
%%   \centering
%%   \includegraphics[width=0.50\textwidth]{path_without_extension}
%%   \caption{Caption below the figure!}\label{fig:example}
%% \end{figure}
%% 
%% \cite{ref}  - Make a citation that is in parenthesis
%% \citeN{ref} - Make a citation out of parenthesis (for noun usage)
%% \shortcite{ref}  - Make a citation for references with >3 authors.
%% \shortciteN{ref} - Make a noun-form citation for references with >3 authors.
%% 
%% 
%% 5 - 12 page paper mandated
%% 150 word abstract
%% 5 keywords maximum

\section*{Abstract}

An algorithm for computing monotone order six piecewise polynomial
interpolants is proposed. Algebraic constraints for enforcing
monotonicity are provided that align with quintic monotonicity
theory. The algorithm is implemented, tested, and applied to several
sample problems to demonstrate the improved accuracy of monotone
quintic spline interpolants compared to that for existing monotone cubic spline
interpolants.

\textbf{Keywords:} monotone, quintic spline, Hermite interpolation, sixth order polynomial.

\section{Introduction and Motivation}
\label{sec:introduction}

%% \begin{figure}[htb]
%%   \centering
%%   \includegraphics[width=0.50\textwidth]{demo_projection}
%%   \caption{These are the projection mechanisms that make a cubic polynomial piece monotone.}\label{fig:projection}
%% \end{figure}

%% The ability to interpolate or regress a function with continuous value and derivatives has proven fundamentally useful. In many applications, there are known shapes or characteristics that are expected of approximations. For example, a cumulative distribution function in statistics cannot decrease, the entropy of a closed thermodynamic system can never increase, an electric motor should not suddenly change direction when slowing down or speeding up. In each of these cases it is important that any constructed spline approximation is monotone.

Many domains of science rely on smooth approximations to real-valued functions over a closed interval. Piecewise polynomial functions (splines) provide the smooth approximations for animation in graphics \cite{herman2015techniques,quint2003scalable}, aesthetic structural support in architecture \cite{brennan2019measure}, efficient aerodynamic surfaces in automotive and aerospace engineering \cite{brennan2019measure}, prolonged effective operation of electric motors \shortcite{berglund2009planning}, and accurate nonparametric approximations in statistics \cite{knott2012interpolating}. While polynomial interpolants or regressors apply broadly, splines are often a good choice because they can approximate globally complex functions while minimizing the local complexity of an approximation.

It is often the case that the true underlying function or phenomenon being modeled has known properties e.g., convexity, positivity, various levels of continuity, or monotonicity. Given a reasonable amount of data, it quickly becomes difficult to achieve desirable properties in a single polynomial function. In general, the maintenance of function properties through interpolation/regression is referred to as \textit{shape preserving} \cite{fritsch1980monotone,gregory1985shape}. The specific shapes this work will achieve in approximations are monotonicity and $C^2$ continuity. These properties are chiefly important to the approximation of cumulative distribution functions and subsequently the effective generation of random numbers from a specified distribution.

In statistics especially, the construction of a monotone interpolating spline that is $C^2$ continuous is meaningfully useful \cite{ramsay1988monotone}. A function with these properties could approximate a cumulative distribution function to a high level of accuracy with relatively few intervals. A twice continuously differentiable approximation to a cumulative distribution function (CDF) would also produce a corresponding probability density function (PDF) that is continuously differentiable, which is a property many standard parametric distributions maintain.

The currently available software for monotone piecewise polynomial interpolation includes quadratic \cite{he1998monotone}, cubic \cite{fritsch1980monotone}, and (with limited application) quartic \cite{wang2004rational,piah2011improved,yao2018unconditionally} cases. In addition, a statistical method for bootstrapping the construction of an arbitrarily smooth monotone fit exists \cite{leitenstorfer2006generalized}, but the method does not take advantage of known analytic properties related to quintic polynomials. Theory has been provided for the quintic case \cite{ulrich1994positivity,hess1994positive}, however this theory has not yet been used to construct a monotone quintic spline interpolation routine. Recent work suggests that the lack of quintic software may be due to a general unawareness of the theory \shortcite{xie2018semiparametric}.

\begin{figure}
  \centering
  \includegraphics[width=.9\textwidth]{spline_demonstration}
  \caption{Example polynomials that interpolate function values at the ends of the interval $[0,1]$. The first only interpolates the function values $g_2(0) = 1$ and $g_2(1) = 0$, making it the order two polynomial $g_2(x) = 1 - x$. For the second plot $g_4(x) = 8x^3 - 12x^2 + 4x + 1/2$, which is order four and interpolates the values $g_4(0) = 1/2$, $g_4'(0) = 4$, $g_4(1) = 1/2$, $g_4'(1) = 4$. Finally the third plot shows the order six polynomial $g_6(x) = - 64x^5 + 160x^4 - 140x^3 + 50x^2 - 7x + 1$ interpolating the function values $g_6(0) = 1$, $g_6'(0) = -7$, $g_6''(0) = 100$, $g_6(1) = 0$, $g_6'(1) = -7$, $g_6''(1) = -100$. Notice that interpolating the same fixed number of function values at each endpoint will always result in an even order interpolating polynomial.
  }\label{fig:spline_demonstration}
\end{figure}

The importance of piecewise quintic interpolation over lower order approximations can be simply demonstrated. In general, the order of a polynomial determines the number of function values it can interpolate, and the growth rate of error away from the interpolated function values. As demonstrated in Figure \ref{fig:spline_demonstration}, it can be seen that matching a value at either end of the interval requires an order two (linear) approximation and each additional derivative at the ends of the interval raises the necessary polynomial order by two. The body of this work is composed of a novel algorithm for enforcing monotonicity on quintic polynomial pieces, then extending that solution to work on quintic splines.

%% Software exists that does monotone piecewise quintic regression \cite{}, but the available routines do not computationally take advantage of the quintic monotonicity theory. ???

The major contribution of this work is an algorithm for constructing monotone quintic interpolating splines that utilizes existing quintic monotonicity theory. The remainder of this paper is structured as follows: Section \ref{sec:monotone-cubic} summarizes the existing monotone cubic spline interpolation methodology, Section \ref{sec:monotone-quintic} presents an algorithm for constructing monotone quintic spline interpolants, Section \ref{sec:results} offers experiments and results with cubic and quintic methods, Sections \ref{sec:discussion} and \ref{sec:conclusion} discuss results and conclude.

\subsection{Computing a Monotone Cubic Interpolant}
\label{sec:monotone-cubic}

The current state-of-the-art monotone interpolating spline with a mathematical software implementation is piecewise cubic, continuously differentiable, and was first proposed in \cite{fritsch1980monotone} then expanded upon in \cite{carlson1985monotone}. Let $\pi: x_0 = k_1 < k_2 < \cdots < k_n = x_1$ be a partition of the interval $[x_0,x_1]$. Let $f: \mathbb{R} \rightarrow \mathbb{R}$ be $C^2$, and $\bigl\{f(k_i)\bigr\}_{i=1}^n$ and $\bigl\{\Delta_i\bigr\}_{i=1}^n$ be given sets of function and derivative values at the partition points for a monotone function $f$. Either $f(k_i) \leq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\ge0$, $i=1$, $\ldots$, $n$, or $f(k_i) \geq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\le0$, $i=1$, $\ldots$, $n$. Let $\hat f$ be a piecewise cubic polynomial defined in each subinterval $I_i = [k_i, k_{i+1}]$ by

\begin{align*}
h_i &= k_{i+1} - k_{i}, \quad
u(t) = 3t^2 - 2t^3, \quad
p(t) = t^3 - t^2,\\
\hat f(x) &= f(k_i) u\big((k_{i+1} - x) / h_i\big) +
f(k_{i+1}) u\big((x - k_i) / h_i\big) \\
&{} - h_i\Delta_i p\big((k_{i+1}-x)/h_i\big) + h_i\Delta_{i+1}
p\big((x-k_i)/h_i\big).
\end{align*}

Notice that a trivially monotone spline results when $\Delta_i = 0$, for $i = 1$, $\ldots$, $n$. However, such a spline has too many \textit{wiggles} for most applications. \citeN{carlson1985monotone} show that simple conditions on the derivative values can guarantee monotonicity, and that these conditions can be enforced in a way that ensures modifications on one interval will not break the monotonicity of cubic polynomials over any neighboring intervals. Consider the terms $\alpha = \big(\Delta_i (k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$ and $\beta = \big(\Delta_{i+1}(k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$. If $f(k_i) = f(k_{i+1})$, take $\Delta_i = \Delta_{i+1} =0$ and $\alpha =\beta =1$. Monotonicity of a cubic polynomial over a subinterval can be maintained by ensuring that $\alpha$ and $\beta$ reside in any of the regions depicted in Figure \ref{fig:projection}.

\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{cubic_projection_demonstration}
  \caption{These are the feasible regions of monotonicity for cubic splines and the projections that make a cubic polynomial piece monotone. The regions themselves are numbered 1--4 corresponding to their original description in Fritsch and Carlson (1980). One point is projected onto each region as a demonstration.}\label{fig:projection}
\end{figure}

The actual region of monotonicity for a cubic polynomial is larger, but projection of $(\alpha, \beta)$ into one of these regions ensures that monotonicity will be achieved and not violated for neighboring regions. The user must decide which region is most appropriate for the projections based on the application, Fritsch and Carlson recommend using Region 2.

While the cubic monotonicity case affords such a concise solution, the region of monotonicity is not so simple in the quintic case. In the next section, an algorithm for performing a projection similar to those for cubic polynomials is proposed.

\section{Monotone Quintic Interpolation}
\label{sec:monotone-quintic}

The following section is composed of three algorithms that together are used to construct a monotone quintic spline interpolant. Without loss of generality, the algorithms will only consider the monotone increasing (nondecreasing) case. The monotone decreasing (nonincreasing) case is handled similarly. Algorithm \ref{alg:check-monotone} checks monotonicity,
Algorithm \ref{alg:make-monotone} enforces monotonicity on an order six
polynomial piece of the quintic spline, and Algorithm
\ref{alg:monotone-spline} uses the previous two algorithms to enforce
monotonicity for the entire quintic spline.

In pseudocode, a function \texttt{binary\_search($g,$ $a,$ $b$)} is used, where $a,b\in S\subset \mathbb{R}^p$ for convex $S$, $g: S \rightarrow \{$\texttt{FALSE}, \texttt{TRUE}$\}$ is a right continuous Boolean function, $g(b) = \hbox{\texttt{TRUE}}$, and for $\mu \in [0,1]$, $g\big((1-\mu)a + \mu b\big) =\hbox{\texttt{TRUE}} \Longrightarrow g\big((1-\nu)a + \nu b\big) =\hbox{\texttt{TRUE}}$ for $\mu \le \nu\le1$. The search returns $\big((1-c)a + c b\big)$ for the smallest $c \in [0,1]$ such that $g\big((1-c)a + c b\big) = \hbox{\texttt{TRUE}}$.

These algorithms make use of first and second derivatives of the approximated function. In the case that the first and second derivative information is not provided along with function values, the derivatives are estimated with finite differences of the function values. The final quintic spline is represented as a piecewise polynomial using the Newton form for each polynomial piece, and ultimately converted to a $B$-spline representation for evaluation.

\algorithm
    {is\_monotone$\big(x_0, x_1, f \big)$}
    {where $x_0$, $x_1 \in \mathbb{R}$, $x_0 < x_1$, and $f$ is an order six polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$,
  $Df(x_1)$, $D^2f(x_1)$. Returns \texttt{TRUE} if $f$ is monotone increasing on $[x_0,x_1]$.}
    {\label{alg:check-monotone}
  \item if $\big(f(x_0) = f(x_1)\big)$ and not $\big( 0 = Df(x_0) = Df(x_1) = D^2f(x_0) = D^2f(x_1) \big)$
    \indented{return FALSE}
    end if
  \item if $\big(Df(x_0) < 0$ or $Df(x_1) < 0\big)$
    \indented{return FALSE}
    end if
    \comment{The necessity of these first two steps follows directly from the fact that $f$ is $C^2$. The next case is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \citeN{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$. Step \ref{alg1-alpha} checks for the necessary condition that $\alpha \ge 0$, Step \ref{alg1-beta} checks $\beta \geq \alpha$, and Step \ref{alg1-gamma} checks $\gamma \geq \delta$, all from \citeN{schmidt1988positivity}. If all necessary conditions are met, then the order six piece is monotone and Step \ref{alg1-simplified-true} concludes this check.}
  \item if $\big(Df(x_0) = 0$ or $Df(x_1) = 0\big)$
  \item \indented{$w := x_0 - x_1$\\$v := f(x_0) - f(x_1)$}
  \item \indented{if $\big(D^2f(x_1) > {-4}Df(x_1) / w \big)$ return FALSE}
    \label{alg1-alpha}
  \item \indented{if $\big(D^2f(x_1) < (3w D^2f(x_0) - 24 Df(x_0) - 32
    Df(x_1) + 60v/w) / (5w) \big)$ return FALSE} \label{alg1-beta}
  \item \indented{if $\big(D^2f(x_0) < 3 Df(x_0) / w \big)$ return FALSE}
    \label{alg1-gamma}
  \item \indented{return TRUE} \label{alg1-simplified-true}
    end if
    \comment{The following code considers the remaining case where $Df(x_0) \neq 0$ and $Df(x_1) \neq 0$.}
  \item $\displaystyle A := Df(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
    \\ $\displaystyle B := Df(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
    \comment{The variables $A$ and $B$ correspond directly to the theoretical foundation for positive quartic polynomials established in \citeN{ulrich1994positivity}, first defined after Equation (18).}
  \item $\displaystyle \gamma_0 := 4 \frac{Df(x_0)}{Df(x_1)} (B/A)^{3/4}$
    \\ $\displaystyle \gamma_1 := \frac{x_1 - x_0}{Df(x_1)} (B/A)^{3/4}$
    \\ $\alpha_0 := 4 (B/A)^{1/4}$
    \\ $\displaystyle \alpha_1 := -\frac{x_1 - x_0}{Df(x_1)} (B/A)^{1/4}$
    \\ $\displaystyle \beta_0 := 30 - \frac{12 \big(Df(x_0) + Df(x_1)\big) (x_1 - x_0)}{\big(f(x_1) - f(x_0)\big) \sqrt{A}\sqrt{B}}$
    \\ $\displaystyle \beta_1 := \frac{-3 (x_1 - x_0)^2}{2 \big(f(x_1) - f(x_0)\big) \sqrt{A} \sqrt{B}} $ \label{alg1-linearized-abg}
    \comment{The $\gamma$, $\alpha$, and $\beta$ terms with subscripts $0$ and $1$ are algebraic reductions of the simplified conditions for satisfying Theorem 2 in Equation (16) of \citeN{ulrich1994positivity}. These terms with subscripts $0$ and $1$ make the computation of $\alpha$, $\beta$, and $\gamma$ functions of the second derivative terms, as seen in Step \ref{alg1-abg} below.}
  \item $\gamma := \gamma_0 + \gamma_1 D^2f(x_0)$
    \\ $\alpha := \alpha_0 + \alpha_1 D^2f(x_1)$
    \\ $\beta := \beta_0 + \beta_1 \big(D^2f(x_0) - D^2f(x_1)\big)$
    \label{alg1-abg}
  \item if $(\beta \leq 6)$ return $\big(\alpha > - (\beta + 2) / 2\big)$
    \\ else $\qquad$ return $\big(\gamma > -2 \sqrt{\beta - 2}\big)$
    \\ end if \label{alg1-last}
}
%% ----------------------------------------------------------------------

The reason for structuring the $\alpha$, $\beta$, and $\gamma$ computations in terms of the second derivative of $f$ (seen in Step \ref{alg1-abg} of Algorithm \ref{alg:check-monotone}) will become more apparent later. The next problem to consider is that of making a nonmonotone order six polynomial piece into a monotone one by modifying its first and second derivative values at the ends of an interval. Note that the actual value of the function at the ends of the interval is not modified, as the resulting polynomial needs to interpolate.

%% ----------------------------------------------------------------------
\algorithm
    {make\_monotone$\big (x_0$,$x_1$,$f \big)$}
    {where $x_0,\ x_1 \in \mathbb{R}$, $x_0 < x_1$, and $f$ is an order six polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$, $Df(x_1)$, $D^2f(x_1)$. Returns $f$ monotone on $[x_0,x_1]$.}
    {\label{alg:make-monotone}
    \item if $\big(f(x_1) = f(x_0)\big)$
      \indented{$Df(x_0) := Df(x_1) := D^2f(x_0) := D^2f(x_1) := 0$}
      \indented{return}
      end if
    \item $Df(x_0) :=\ $median$\displaystyle \bigg(0,\ Df(x_0),\ 14 \frac{f(x_1) - f(x_0)}{x_1 - x_0} \bigg)$
      \\ $Df(x_1) :=\ $median$\displaystyle \bigg(0,\ Df(x_1),\ 14 \frac{f(x_1) - f(x_0)}{x_1 - x_0}\bigg)$ \label{alg2-init-df}
      \comment{The selection of values in Step \ref{alg2-init-df} for $Df(x_0)$ and $Df(x_1)$ is suggested by \citeN{ulrich1994positivity} and also by \citeN{huynh1993accurate}. These rules quickly enforce upper and lower bounds on derivative values to ensure quintic monotonicity is obtainable.}
      \comment{The next case that will be covered is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \citeN{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$. Steps \ref{alg2-dx-conditions} and \ref{alg2-rescale-dx} ensure that the following bounds on $D^2f(x_0)$ are compatible while Step \ref{alg2-ddx0-upper} ensures that the conditions on $D^2f(x_1)$ are compatible. To conclude the simplified case, Step \ref{alg2-gamma} first enforces $\gamma \geq \delta$, then Steps \ref{alg2-alpha} and \ref{alg2-beta} enforce $\alpha \geq 0$ and $\beta \geq \alpha$ respectively. Combined, these three conditions derived from Equation (2.18) in \citeN{schmidt1988positivity} guarantee monotonicity of this polynomial piece.}
    \item if $\big(Df(x_0) = 0$ or $Df(x_1) = 0\big)$
    \item \indented{$w := x_0 - x_1$\\$v := f(x_0) - f(x_1)$}
    \item \indented{if $\big(5 Df(x_0) + 4Df(x_1) > 20v / w)$} \label{alg2-dx-conditions}
    \item \indented{\indented{$Df(x_0) := Df(x_0) \text{ max}\displaystyle \bigg(0, \frac{20v}{w \big(5 Df(x_0) + 4 Df(x_1)\big)}\bigg) $}}
      \indented{\indented{$Df(x_1) := Df(x_1) \text{ max}\displaystyle \bigg(0, \frac{20v}{w \big(5 Df(x_0) + 4 Df(x_1)\big)}\bigg) $}} \label{alg2-rescale-dx}
      \indented{end if}
    \item \indented{$D^2f(x_0) :=\ $min$\displaystyle \bigg(D^2f(x_0), \frac{4 (2 Df(x_0)+Df(x_1))+20 v/w}{w} \bigg)$} \label{alg2-ddx0-upper}
    \item  \indented{$D^2f(x_0) :=\ $max$\big(D^2f(x_0), 3 Df(x_0) / w \big)$} \label{alg2-gamma}
    \item \indented{$D^2f(x_1) :=\ $min$\big(D^2f(x_1), {-4}Df(x_1) / w \big)$} \label{alg2-alpha}
    \item  \indented{$D^2f(x_1) :=\ $max$\displaystyle \bigg(D^2f(x_1), \frac{3w D^2f(x_0) - 24 Df(x_0) - 32 Df(x_1) + 60v/w}{5w} \bigg)$} \label{alg2-beta}
      \indented{return}
      end if
      \comment{The following code considers the case where $Df(x_0) \neq 0$ and $Df(x_1) \neq 0$.}
    \item $\displaystyle A := Df(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
      \\ $\displaystyle B := Df(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
    \item if $\big($max$(A,B) > 6\big)$
      \indented{$Df(x_0) := 6 Df(x_0) \big/ $max$(A,B)$}
      \indented{$Df(x_1) := 6 Df(x_1) \big/ $max$(A,B)$}
      end if
      \comment{This ensures that $(A,B)$ remains within a viable region of monotonicity (satisfying Theorem 4, seen in Fig. 6 of \citeN{ulrich1994positivity}).}
    \item $\eta := \big(D^2f(x_0), D^2f(x_1)\big)$
      \\ $\displaystyle \eta_0 := \left({-\frac{\sqrt{A}}{4}}\big(7\sqrt{A} + 3\sqrt{B}\big),\ \frac{\sqrt{B}}{4}\big(3\sqrt{A} + 7\sqrt{B}\big) \right)$
    \item $\bigl(D^2f(x_0)$, $D^2f(x_1)\bigr) :=$ binary\_search$\bigl(g,
       \eta,\eta_0\bigr)$
      \comment{Here $g(z) =$\texttt{ is\_monotone}$(x_0,x_1,f)$ where $x_0$, $x_1$, $f(x_0)$, $f(x_1)$, $Df(x_0)$, $Df(x_1)$ are fixed, and the variable $z=\bigl(D^2f(x_0)$,$D^2f(x_1)\bigr)$. This binary search identifies feasible values of $D^2f(x_0)$ and $D^2f(x_1)$ that are nearest to the current values $\eta$. This search is valid because $\eta_0$ is guaranteed to produce a monotone function, which can be seen in Equation (23) of \citeN{ulrich1994positivity}.}
    return
}
%% ----------------------------------------------------------------------

>>>>

\documentclass[12pt,letterpaper]{article}
\usepackage{fullpage}  % Make the margins smaller
\usepackage{parskip}   % Don't use indents in front of paragraphs
\usepackage{xcolor}    % Color package for coloring text.
\usepackage{fontenc}   % Use a font package.
\usepackage{libertine} % Pick the font.
\usepackage{hyperref} % Links 
\usepackage{amsfonts}  % Math font "mathbb"

%% Use a custom `algorithm` command for this paper.
\input{lux-algorithm}
\renewcommand*{\ttdefault}{courier}

\newcommand{\TODO}[1]{{\color{red} #1}}

\medskip

\title{Piecewise Monotone Quintic Spline Interpolants}
\author{Thomas C. H. Lux}
\date{}

\begin{document}
\thispagestyle{empty}
\begin{center}
    {\LARGE \textbf{Piecewise Monotone Quintic Spline Interpolants}} \hfill
    %% \href{https://tchlux.info}{\textbf{tchlux.info}} \hfill
    %% \textbf{804 317 8339} \hfill
    %% \href{thomas.ch.lux@gmail.com}{\textbf{thomas.ch.lux@gmail.com}}
    \\ \medskip\href{tchlux@vt.edu}{Thomas C.H. Lux}
    %% \rule{\textwidth}{1.5pt}
    \bigskip
\end{center}

%% \begin{tabular}{@{}l r@{}}
%%     \begin{tabular}{@{}l@{}}
%%       %% Hiring Committee \\
%%       Business \\
%%       City, State \\
%%       \today
%%     \end{tabular} &
%%     %% \hfill IMAGE
%% \end{tabular}

\medskip

\algorithm
    {is\_monotone$\big(x_0, x_1, f \big)$}
    {where $x_0$, $x_1 \in \mathbb{R}$, $x_0 < x_1$, and $f$ is an order six polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$,
  $Df(x_1)$, $D^2f(x_1)$. Returns \texttt{TRUE} if $f$ is monotone on $[x_0,x_1]$.}
    {\label{alg:check-monotone}
  \item if $\big(f(x_0) = f(x_1)\big)$ and not $\big( 0 = Df(x_0) = Df(x_1) = D^2f(x_0) = D^2f(x_1) \big)$
    \indented{return FALSE}
    else if $\big(f(x_0) < f(x_1)\big)$ ; $s = 1$
    \\else ; $s = -1$
    \\end if
  \item if $\big(Df(x_0) < 0$ or $Df(x_1) < 0\big)$
    \indented{return FALSE}
    end if
    \comment{The necessity of these first two steps follows directly from the fact that $f$ is $C^2$. The next case is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \cite{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$. Step \ref{alg1-alpha} checks for the necessary condition that $\alpha \ge 0$, Step \ref{alg1-beta} checks $\beta \geq \alpha$, and Step \ref{alg1-gamma} checks $\gamma \geq \delta$, all from \cite{schmidt1988positivity}. If all necessary conditions are met, then the order six piece is monotone and Step \ref{alg1-simplified-true} concludes this check.}
  \item if $\big(Df(x_0) = 0$ or $Df(x_1) = 0\big)$
  \item \indented{$w := x_0 - x_1$\\$v := f(x_0) - f(x_1)$}
  \item \indented{if $\big(D^2f(x_1) > {-4}Df(x_1) / w \big)$ return FALSE}
    \label{alg1-alpha}
  \item \indented{if $\big(D^2f(x_1) < (3w D^2f(x_0) - 24 Df(x_0) - 32
    Df(x_1) + 60v/w) / (5w) \big)$ return FALSE} \label{alg1-beta}
  \item \indented{if $\big(D^2f(x_0) < 3 Df(x_0) / w \big)$ return FALSE}
    \label{alg1-gamma}
  \item \indented{return TRUE} \label{alg1-simplified-true}
    end if
    \comment{The following code considers the remaining case where $Df(x_0) \neq 0$ and $Df(x_1) \neq 0$.}
  \item $\displaystyle A := Df(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
    \\ $\displaystyle B := Df(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
    \comment{The variables $A$ and $B$ correspond directly to the theoretical foundation for positive quartic polynomials established in \cite{ulrich1994positivity}, first defined after Equation (18).}
  \item $\displaystyle \gamma_0 := 4 \frac{Df(x_0)}{Df(x_1)} (B/A)^{3/4}$
    \\ $\displaystyle \gamma_1 := \frac{x_1 - x_0}{Df(x_1)} (B/A)^{3/4}$
    \\ $\alpha_0 := 4 (B/A)^{1/4}$
    \\ $\displaystyle \alpha_1 := -\frac{x_1 - x_0}{Df(x_1)} (B/A)^{1/4}$
    \\ $\displaystyle \beta_0 := 30 - \frac{12 \big(Df(x_0) + Df(x_1)\big) (x_1 - x_0)}{\big(f(x_1) - f(x_0)\big) \sqrt{A}\sqrt{B}}$
    \\ $\displaystyle \beta_1 := \frac{-3 (x_1 - x_0)^2}{2 \big(f(x_1) - f(x_0)\big) \sqrt{A} \sqrt{B}} $ \label{alg1-linearized-abg}
    \comment{The $\gamma$, $\alpha$, and $\beta$ terms with subscripts $0$ and $1$ are algebraic reductions of the simplified conditions for satisfying Theorem 2 in Equation (16) of \cite{ulrich1994positivity}. These terms with subscripts $0$ and $1$ make the computation of $\alpha$, $\beta$, and $\gamma$ functions of the second derivative terms, as seen in Step \ref{alg1-abg} below.}
  \item $\gamma := \gamma_0 + \gamma_1 D^2f(x_0)$
    \\ $\alpha := \alpha_0 + \alpha_1 D^2f(x_1)$
    \\ $\beta := \beta_0 + \beta_1 \big(D^2f(x_0) - D^2f(x_1)\big)$
    \label{alg1-abg}
  \item if $(\beta \leq 6)$ return $\big(\alpha > - (\beta + 2) / 2\big)$
    \\ else $\qquad$ return $\big(\gamma > -2 \sqrt{\beta - 2}\big)$
    \\ end if \label{alg1-last}
}

\medskip

\section*{Comments}

This is a final section discussing some of the finer points.

%% \medskip

%% With respect,\\
%% \bigskip\\
%% \bigskip\\
%% Thomas Lux
%% \vfill

\bibliographystyle{unsrt}
\bibliography{pmqsi}
